---
title: "Notes on Québec AQRP preprocessing"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    highlight: zenburn
    theme: sandstone
    toc: yes
---

```{r, setup, include=FALSE, purl=FALSE}
# Set code chunk options
knitr::opts_chunk$set(
  # Suppress warnings/messages
  warning = FALSE, 
  message = FALSE,
  # Center plots
  fig.align = 'center',
  fig.width = 8.5
)

# Hide this workflow dependency for greater simplicity
tar_load(study_area)
```

# 0. Define targets workflow dependency
```{r}
targets::tar_load(aqrp_roads)
targets::tar_load(building_footprint)
targets::tar_load(quebec_roads)
```


# 1. Visualize Québec's AQRP road network dataset

The AQRéseau+ (AQRP) dataset is publicly available here:
`https://www.donneesquebec.ca/recherche/dataset/adresses-quebec/resource/a23ac6ee-2912-47ba-bec5-23fb9ddc3c7d`.

Information on the metadata can be found here:
`https://www.donneesquebec.ca/recherche/dataset/adresses-quebec/resource/48161699-adba-4dcc-b5b8-27f699ae2f2a`.

The AQRP contains provincial/municipal road network data from the ministère des Transports et de la Mobilité durable (MTMD) and the ministère des Affaires municipales et de l’Habitation (MAMH) and contains logging road network data from the ministère des Ressources naturelles et des Forêts.

```{r}
# Bounding box object
aqrp_bbox <- sf::st_bbox(aqrp_roads)

# Admin boundaries
statcan_qc_on <- sf::read_sf("data/study_area/statcan_admin_bounds.shp") %>%
 # Filter QC and ON
 dplyr::filter(PRUID %in% c(24, 35))

# Quick plot
ggplot2::ggplot() +
 # Add Québec/Ontario polygon
 ggplot2::geom_sf(data = statcan_qc_on) +
 # Add caribou range
 ggplot2::geom_sf(data = study_area, fill = "forestgreen", alpha = 0.7) +
 # Add AQRP roads
 ggplot2::geom_sf(data = aqrp_roads) +
 # Add coords
 ggplot2::coord_sf(
   xlim = c(aqrp_bbox[1], aqrp_bbox[3]),
   ylim = c(aqrp_bbox[2], aqrp_bbox[4])
 ) +
 theme_caribou +
 NULL
```

# 2. Exploring road attributes in study area

## 2.1 Road surface type

In the AQRP dataset, the `Etat_Rev` attribute describes the surface of the road.  
```{r warning=FALSE}
ggplot2::ggplot(data = aqrp_roads) +
  ggplot2::geom_histogram(
    mapping = ggplot2::aes(x = etat_rev),
    stat = "count"
  ) +
  ggplot2::labs(
    x = "Surface type",
    y = "Count"
  ) +
  theme_caribou
```

We will first recategorize `Etat_Rev` into `paved` and `unpaved` under the new attribute `surface`. While the majority of all roads in the study area are `Non revêtue` (unpaved), note the presence of `Autre` (other), `ND` (non-determined), and `NA`s which we will include as `unknown` for now. We will later reclassify `unknown`s into `paved` and `unpaved` using various criteria.
```{r warning=FALSE}
aqrp_roads <- aqrp_roads %>% 
  dplyr::mutate(
    surface = dplyr::case_when(
      etat_rev == "Non revêtue" ~ "unpaved",
      etat_rev == "Revêtue" ~ "paved",
      # NA as unknown
      is.na(etat_rev) ~ "unknown",
      # Autre and ND as unknown
      TRUE ~ "unknown"
    )
  )
```

## 2.2 Forest road classification

In the AQRP dataset, the `Cls_CheFor` attribute describes road classifications for forestry roads.
```{r warning=FALSE}
ggplot2::ggplot(data = aqrp_roads) +
  ggplot2::geom_histogram(
    mapping = ggplot2::aes(x = cls_che_for),
    stat = "count"
  ) +
  ggplot2::labs(
    x = "Forest road classification",
    y = "Count"
  ) +
  theme_caribou
```

Note the presence of `NA`s which will toss into the `IN` (Inconnu, unknown) placeholder to avoid filtering mistakes downstream.
```{r}
aqrp_roads <- aqrp_roads %>% 
  # Replace NAs prior to filtering to avoid dropping
  tidyr::replace_na(list(cls_che_for = "IN"))
```

Let's inspect the proportion of the total length of each of these categories that are `unpaved`, `paved`, or `unknown`.
```{r fig.height=6, warning=FALSE, message=FALSE}
cls_che_for_proportions <- aqrp_roads %>% 
  # Compute length of each road segment then drop geometries
  dplyr::mutate(length = sf::st_length(geometry)) %>%
  sf::st_drop_geometry() %>% 
  # Group by Cls_CheFor and compute total length of road in each class
  dplyr::group_by(cls_che_for) %>% 
  dplyr::mutate(cls_che_for_length_total = sum(length)) %>% 
  # Summarize total length for each surface type within each road class
  dplyr::group_by(cls_che_for, surface) %>% 
  dplyr::summarise(total_length_surface = sum(length)) %>%
  # Compute total length of each class, then proportion of surfaces in-class
  dplyr::mutate(
    total_length_cls_che_for = sum(total_length_surface),
    proportion = as.numeric(total_length_surface/total_length_cls_che_for)
  ) %>%
  # Clean up
  dplyr::ungroup()

# Visualize proportions of paved/unpaved/unknown surface in each road class
ggplot2::ggplot(cls_che_for_proportions, ggplot2::aes(cls_che_for, proportion)) +
  ggplot2::geom_bar(
    mapping = ggplot2::aes(fill = surface),
    position = "stack", stat = "identity"
  ) +
  ggplot2::scale_fill_viridis_d(direction = -1) +
  ggplot2::labs(
    x = "Forest road classification",
    y = "Proportion",
    fill = "Surface type"
  ) +
  ggplot2::coord_flip() +
  theme_caribou +
  ggplot2::theme(legend.position = "top")
```

Note that for all classes except `NF`, roads are overwhelmingly `unpaved`. The `paved` portion of all other classes are highly suspect. Closer inspection reveal that these are mainly small bridges on logging roads which we address later.

## 2.3 Road classification attribute

In the AQRP dataset, the `Cls_Rte` attribute describes the broader road classification.
```{r fig.height=6, warning=FALSE}
ggplot2::ggplot(data = aqrp_roads) +
  ggplot2::geom_histogram(
    mapping = ggplot2::aes(x = cls_rte),
    stat = "count"
  ) +
  ggplot2::labs(x = "Road classification", y = "Count") +
  ggplot2::scale_y_log10() +
  theme_caribou +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5)
  )
```

Note the large number of `NA`s which will need to be replaced with the `IN` (Inconnu) placeholder to avoid filtering mistakes downstream. Also note the `Laison maritime` (ferry routes), which will need to be removed prior to analysis.
```{r}
aqrp_roads <- aqrp_roads %>% 
  # Replace NAs prior to filtering to avoid dropping
  tidyr::replace_na(list(cls_rte = "IN")) %>% 
  # Drop ferry routes
  dplyr::filter(cls_rte != "Liaison maritime")
```

Let's inspect the prevalence of each class and the proportion of the total length of each of these classes that are `unpaved`, `paved`, and `unknown` in the subset of non-forest roads identified in Section 2.2.
```{r fig.height=6, warning=FALSE, message=FALSE}
# Get subset
non_forest_roads <- aqrp_roads %>% 
  # Non-forest roads (77,422 roads of 664,540)
  dplyr::filter(cls_che_for == "NF")

# Compute proportions
cls_road_proportions <- non_forest_roads %>% 
  # Compute length of each road segment then drop geometries
  dplyr::mutate(length = sf::st_length(geometry)) %>%
  sf::st_drop_geometry() %>% 
  # Group by Cls_Rte and compute total length of road in each class
  dplyr::group_by(cls_rte) %>% 
  dplyr::mutate(cls_rte_length_total = sum(length)) %>% 
  # Summarize total length for each surface type within each road class
  dplyr::group_by(cls_rte, surface) %>% 
  dplyr::summarise(total_length_surface = sum(length)) %>%
  # Compute total length of each class, then proportion of surfaces in-class
  dplyr::mutate(
    total_length_class = sum(total_length_surface),
    proportion = as.numeric(total_length_surface/total_length_class)
  ) %>%
  # Clean up
  dplyr::ungroup()

# Subset road class counts
ggplot2::ggplot(
  data = tidyr::complete(non_forest_roads, cls_rte, surface),
  mapping = ggplot2::aes(cls_rte)
) +
  ggplot2::geom_bar(
    mapping = ggplot2::aes(fill = surface),
    position = "dodge", stat = "count"
  ) +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::labs(
    x = "Road classification",
    y = "Count",
    fill = "Surface type"
  ) +
  ggplot2::coord_flip() +
  
  theme_caribou +
  ggplot2::theme(legend.position = "top")


# Visualize proportions of paved/unpaved/unknown surface in each road class
ggplot2::ggplot(cls_road_proportions, ggplot2::aes(cls_rte, proportion)) +
  ggplot2::geom_bar(
    mapping = ggplot2::aes(fill = surface),
    position = "stack", stat = "identity"
  ) +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::labs(
    x = "Road classification",
    y = "Within-class proportion",
    fill = "Surface type"
  ) +
  ggplot2::coord_flip() +
  theme_caribou +
  ggplot2::theme(legend.position = "top")
```
Nothing too conclusive here, but:
* The most abundant road class is `IN` (inconnu, unknown) and is likely to be all `unpaved` surfaces.
* Within the `Nationale`, `Régionale`, `Locale`, `Collectrice de municipale`, `Collectrice de transit`, `Autoroute`, and `Artère` road classes, it is likely that `unknown` surfaces are `paved`. 

## 2.4 Road classification versus building footprint map

When a road surface is unknown, the proximity to built up areas is a great indicator of the presence of pavement. We will use Microsoft's global building footprint available [here](https://github.com/microsoft/GlobalMLBuildingFootprints) and preprocessed to create a 2-km buffer around buildings in the study area using the script `R/get_building_footprint.R`. We will overlay non-forest roads with `unknown` surface over the building footprint. 

```{r}
# Bounding box object
non_forest_roads_bbox <- sf::st_bbox(non_forest_roads)

# Admin boundaries
statcan_qc_on <- sf::read_sf("data/study_area/statcan_admin_bounds.shp") %>%
 # Filter QC and ON
 dplyr::filter(PRUID %in% c(24, 35))

# Quick plot
unknown_surface_plot <- ggplot2::ggplot() +
  # Add Québec/Ontario polygon
  ggplot2::geom_sf(data = statcan_qc_on) +
  # Add global building footprint 2-km buffer
  ggplot2::geom_sf(data = building_footprint, fill = "grey30") +
  # Add nopn-forest roads subset
  ggplot2::geom_sf(
    data = dplyr::filter(non_forest_roads, surface == "unknown"),
    colour = "red"
  ) +
  # Labs
  ggplot2::labs(colour = "Surface type") +
  # Add coords
  ggplot2::coord_sf(
    xlim = c(non_forest_roads_bbox[1], non_forest_roads_bbox[3]),
    ylim = c(non_forest_roads_bbox[2], non_forest_roads_bbox[4])
  ) +
  theme_caribou +
  NULL

# Add interactive map
plotly::ggplotly(unknown_surface_plot)
```

Most non-forest roads with an `unknown` surface type are concentrated more densely populated areas in the southern part of the study area, such as Lac St-Jean and Abitibi. Also highlighted are the long roads for the LG Hydro-Québec complexes in the northern part of the study range.   

Two large points need to be addressed: 

1. The remaining roads with `unknown` surface type, heavily concentrated in non-forest roads, need to be reclassified as `paved` or `unpaved`.
2. There are tiny isolated segments of `paved` road sprinkled throughout logging road networks (often bridges) that need to be reclassified as `unpaved`.

# 3. Reclassification of `unknown`s as paved or unpaved

## 3.1 Forest roads

In forest roads, all `surface` that remain `unknown` will be reclassified as `unpaved`.
```{r}
# Create subset of forest roads
forest_roads_reclass <- aqrp_roads %>% 
  dplyr::filter(cls_che_for != "NF") %>% 
  # Reclassify
  dplyr::mutate(
    surface = dplyr::if_else(
      # If unknown surface
      condition = surface == "unknown",
      # Reclassify as unpaved
      true = "unpaved",
      # Otherwise keep the stated surface type
      false = surface
    )
  )
```

## 3.2 Non-forest roads

In non-forest roads, there are a lot of `unknown`s. If a road is within 2km of a built up area and it's surface is `unknown`, we will reclassify it as `paved`, while anything further than 2km we will reclassify as `unpaved`.
```{r}
# Subset non-forest roads with known surface
nonforest_roads_known_surface <- aqrp_roads %>% 
  dplyr::filter(cls_che_for == "NF", surface != "unknown")

# Reclassify non-forest roads with unknown surface as paved if...
nonforest_roads_reclassed_paved <- aqrp_roads %>% 
  dplyr::filter(cls_che_for == "NF", surface == "unknown") %>% 
  # ...they intersect the building footprint's 2km buffer
  sf::st_filter(building_footprint) %>% 
  dplyr::mutate(surface = "paved")

# Reclassify remaining non-forest roads with unknown surface as unpaved
nonforest_roads_reclassed_unpaved <- aqrp_roads %>% 
  dplyr::filter(cls_che_for == "NF", surface == "unknown") %>% 
  dplyr::filter(!(aqrp_uuid %in% nonforest_roads_reclassed_paved$aqrp_uuid)) %>% 
  dplyr::mutate(surface = "unpaved")

# Bind rows of subsets
nonforest_roads_reclassed <- dplyr::bind_rows(
  nonforest_roads_known_surface,
  nonforest_roads_reclassed_paved,
  nonforest_roads_reclassed_unpaved
)
```

Lets quickly check that we have the desired changes, there should be no more non-forest roads with an `unknown` surface type.
```{r}
# Tally surface types before reclassification
dplyr::filter(aqrp_roads, cls_che_for == "NF") %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(surface) %>%
  tally()

# Tally surface types after reclassification
nonforest_roads_reclassed %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(surface) %>%
  tally()
```

There should not be any roads left with `unknown` surface type.
```{r}
dplyr::bind_rows(nonforest_roads_reclassed, forest_roads_reclass) %>% 
  sf::st_drop_geometry() %>% 
  group_by(surface) %>% 
  tally()
```


# 4. Reclassification of small isolated `paved` segments to `unpaved`

Throughout the study area, small bridges across creeks encountered by logging roads are categorized as `paved` roads, which isn't quite correct. For the purposes of this project we will reclassify these as `unpaved`. This preprocessing step is computation-heavy, so we provide a small reproducible example illustrating the steps of the procedure.

## 4.1 Small area map for reproducible example


```{r warning=FALSE}
# Define bbox of reproducible example
reprex_bbox <- aqrp_roads %>%
  # This bridge is a logging road and not paved
  dplyr::filter(aqrp_uuid == "8eab84e2-b859-4bfd-a184-b738990548c8") %>% 
  sf::st_buffer(500) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() 
  
# Subset roads
reprex_aqrp <- aqrp_roads %>%
  sf::st_intersection(reprex_bbox)

# Plot object
reprex_plot <- ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = reprex_aqrp,
    mapping = ggplot2::aes(colour = surface, linewidth = surface)
  ) +
  ggplot2::scale_linewidth_manual(values = c(1, 0.5)) +
  ggplot2::labs(
    colour = "",
    linewidth = ""
  ) +
  theme_caribou

# Interactive plot
plotly::ggplotly(reprex_plot)
```

## 4.2 Reclassify as `unpaved` using spatial filter
```{r}
# Reduce size of dataset used for spatial filtering
paved_roads_to_check <- dplyr::filter(reprex_aqrp, surface == "paved")
unpaved_roads <- dplyr::filter(reprex_aqrp, surface == "unpaved")

# Reclassify spatially isolated paved segments
paved_roads_reclass <- paved_roads_to_check %>% 
  # Nest by row
  dplyr::group_nest(dplyr::row_number()) %>% 
  # Extract list
  dplyr::pull(data) %>% 
  # Map over each paved road
  purrr::map(
    # Spatial logic filtering function
    .f = ~ {
      # Compute number of paved roads nearby
      n_paved_roads_nearby <- paved_roads_to_check %>%
        # Remove the paved segment of interest from aqrp dataset
        dplyr::filter(aqrp_uuid != .x$aqrp_uuid) %>% 
        # Filter for any intersecting roads within 100 m
        sf::st_filter(sf::st_buffer(.x, 100)) %>%
        # Count how many are paved
        dplyr::filter(surface == "paved") %>%
        nrow()
      
      # If none of the neighbouring roads are paved
      if(n_paved_roads_nearby == 0) {
        # Reclassify the road of interest as unpaved
        dplyr::mutate(.x, surface = "unpaved")
      } else {
        # Else return unchanged road
        .x
      }
    }
  ) %>%
  # Recombine list of sf objects into single table
  purrr::list_rbind() %>%
  # Cast to sf
  sf::st_as_sf() %>% 
  # # Testing
  # dplyr::glimpse() %>% 
  {.}

# Join back to unpaved roads
reprex_aqrp <- dplyr::bind_rows(unpaved_roads, paved_roads_reclass)
```

## 4.3 Visualize the reclassified roads
```{r}
# Plot object
reclass_plot <- ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = reprex_aqrp,
    mapping = ggplot2::aes(colour = surface, linewidth = surface)
  ) +
  ggplot2::scale_linewidth_manual(values = c(0.5, 1),) +
  ggplot2::scale_colour_manual(values = c("#00BFC4","#F8766D")) +
  ggplot2::labs(
    colour = "",
    linewidth = ""
  ) +
  theme_caribou

# Interactive plot
plotly::ggplotly(reclass_plot)
```
# 5. Visualizing final output

All the steps described herein are implemented in a more computationally efficient parallel processing target described in the function `preprocess_quebec_roads()` in the `preprocess_road_networks.R` script. 

We can visualize the final output of that function here.
```{r}
# Quick plot
ggplot2::ggplot() +
  # Add Québec/Ontario polygon
  ggplot2::geom_sf(data = statcan_qc_on) +
  # Add caribou range
  ggplot2::geom_sf(data = study_area, fill = "forestgreen", alpha = 0.2) +
  # Add preprocessed Québec roads
  ggplot2::geom_sf(
    data = quebec_roads %>% 
      dplyr::mutate(
        surface = forcats::as_factor(surface),
        surface = forcats::fct_relevel(surface, "unpaved")
      ),
    mapping = ggplot2::aes(colour = surface)
  ) +
  # Add legend title
  ggplot2::labs(colour = "Road surface") +
  # Add coords
  ggplot2::coord_sf(
    xlim = c(aqrp_bbox[1], aqrp_bbox[3]),
    ylim = c(aqrp_bbox[2], aqrp_bbox[4])
  ) +
  theme_caribou +
  NULL
```


